How has the Amazon been constructed as a problem? Presidential speeches and transnational politics since 1985

### 2023-02-14 Submission/Re-submission/draft/draft_resubmission_shorten.rmd
newest

---
title: How has the Amazon been constructed as a problem? Presidential speeches and transnational politics since 1985
author:
- Livio Silva-Muller
- Henrique Sposito^[Both authors contributed equally to the article. Names are ordered alphabetically. Livio is a PhD Candidate in Sociology and affiliated to the Albert Hirschman Centre on Democracy. Henrique is a PhD Candidate in International Relations and Political Sciences and affiliated to the Centre for International Environmental Studies.]
date: 13 February 2023
output:
  pdf_document: default
  latex_engine: xelatex
  fig_caption: yes
  html_document:
    df_print: paged
fontsize: 12pt
mainfont: Times New Roman
header-includes:
   - \usepackage{floatrow}
   - \floatsetup[figure]{capposition=top}
   - \usepackage{float}
   - \floatplacement{figure}{H}
   - \usepackage[T1]{fontenc}
   - \usepackage[utf8]{inputenc}
   - |
    \providecommand{\keywords}[1]{\textbf{Keywords:} #1}
    \providecommand{\wordcount}[1]{\textbf{Word Count:}#1}
always_allow_html: TRUE
bibliography: references.bib
csl: chicago.csl
abstract: The Amazon is often approached as a transnational environmental problem without a discussion of how key actors see it differently in space and time. We combine theories of problem-representation and transnationalism to analyze where, when, and how the Brazilian Amazon has been constructed as a problem. We develop a framework that explains variation in transnational problem-constructions based on speech setting and problem pressingness. To test our framework, we use supervised machine learning to classify statements about the Amazon in 6240 presidential speeches since 1985. We find that when presidents speak outside (or inside) the Amazon and environmental problems are less (or more) pressing, presidents are likelier to highlight issues of environmental conservation (or economic integration). Furthermore, we show how we are moving from a period when the Amazon as a binary of either environmental or an economic issue has been overcome, to a moment when diverse problem-constructions are deployed.

@@

```{=latex}
\wordcount{ 8765 (including abstract, text, references, and footnotes)}
```

@@

Scholars studying the Brazilian Amazon often identify different governments throughout Brazilian history as proponents of a single view of the region, its people, and forest [see @drummond2006; @padua2012; @franchini2019; @capobianco2019; @pereira2021]. While the military dictatorship (1964-85) and Bolsonaro’s presidency (2019-2022) are, for example, associated with understanding the Amazon as issues of national sovereignty and economic integration, Sarney’s (1985-1990) and Lula’s (2003-2010) presidencies are associated with environmental conservation. Yet, such representations advance the view that governments act consistently towards the Amazon as a specific problem, masking government diversity of in policies and discourse. In this article, we show where, when, and how the Brazilian Amazon has been constructed as a problem in presidential discourses since 1985.

By combining @bacchi1999’s theory of problem representation and @keck1998's theory of transnationalism, we place the way the objects of policies are constructed in discourse at the center of our analysis. We offer a framework that explains variation in discursive problem-constructions depending on speech setting and problem pressingness as actors participate and shape transnational politics. We focus on speeches, rather than policies, because they allow us to capture the diversity of ways the Amazon, as a policy object that demands dedicated attention, has been discursively constructed as a problem. While many analyses of Brazilian environmental discourse and policies intend to dissect its national and international nature, they rarely adopt methodological strategies that distinguish how important actors behave differently in both scenarios [@hochstetler2007]. Our framework and methodology help identify and account for these variations. We show, for instance, that when speaking to farmers in Cuiaba, one of Brazil’s agricultural powerhouses, in a moment when the environmental problem is pressing, it is likely that any president would sidestep environmental issues, as President Lula did in the opening epigraph.

@@

We have four main findings. First, we find that presidents adopt different constructions of the Amazon in different settings according to how they expect to leverage audiences for symbolic and/or  material gains. Second, presidents often focus on alternative problem-constructions when problems become more pressing; whereas they boast about positive outcomes when problems are less pressing. Third, we find that high-profile environmental related events, which receive strong international media attention, increase the importance of the Amazon in presidential speeches. Alternatively, relevant Amazonian-related policies appear to not to attract much attention. Finally, we find that while all presidents deploy multiple problem-constructions across space and over time, we are now in a period when the dictum between understanding the Amazon as an environment or as an economic issue has been overcome, at least discursively. Conceptually, our research shows how focusing on problem-constructions highlight what certain objects of policy are assumed to be, and when they are ignored or privileged, based on the transnational setting of the speech and the pressingness of aligned problems.

@@

# 2. Theory: problem-constructions, transnational settings, and problem pressingness

We divide our theoretical framework in two parts. We start with an explanation of what are problem-constructions and why presidential speeches are an interesting instance to identify them. Next, we argue that problem-constructions can vary depending on the transnational setting of the speech and on the pressingness of its problem.

## 2.1 Problem representation in presidential discourse

Social scientists often describe federal governments as proponents of a cohesive set of policies toward the Amazon [see @drummond2006; @padua2012; @franchini2019; @capobianco2019; @pereira2021]. The 1964 military dictatorship, for example, is associated with securing sovereignty in the region by populating it and integrating it into the national economy [@drummond2006]. Whereas, governments from the late-1980s up to the late-2000s are associated with a turn towards policies that focus on environmental conservation of the Amazon [@padua2012]. The presidencies of Rousseff (2011-2016) and Temer (2016-2018) are connected to the de-prioritizing of environmental conservation policies, while Bolsonaro (2019-2022) with the dismantling of environmental policies [@capobianco2019]. Although these works are important to understand how different governments acted towards the Amazon, they represent governments monolithically by bundling them into cohesive policy periods, or policy cycles, with one all-encompassing understanding of the Amazon.

@@

## 2.2 Transnational settings, pressingness, and variation in problem-constructions

@@

Problem-constructions are choices made by presidents when speaking. If malleable identities across space and time are a defining feature of transnational politics, it follows that presidents' choice of problem-constructions depend on the president's views of an audience at a certain period. Specifically, how pressing a problem is in a certain space and/or at a certain time.

@@

Presidents construct problems and attempt to inform and/or leverage transnational audiences for symbolic or material gains according to their expectations of that audience. We conceptualize spaces in terms of settings. Settings are levels of aggregation for transnational audiences that make sense in light of the object of policy under study. This indicates that how presidents construct policy objects as problems can vary, for instance, within the country — speaking inside the Amazon region versus outside — because presidents identify these audiences as sources of distinct types of support and demand. The same holds true when speaking abroad — in Amazonian countries, non-Amazonian countries, and at international summits — where presidents’ expectations and identities of an audiences could diverge completely from that of domestic actors. In each one of these settings different problems might be more pressing and the president's expectations about what can be leveraged for material and symbolic gains might influence how a problem is implicitly or explicitly constructed.

@@

How presidents construct policy objects can also vary over time because the pressingness of certain problems changes as new information arises. We conceptualize time as the perceived pressingness of a broad and related issue to those surrounding a policy object. This indicates, for instance, that as climate change becomes more pressing and mainstreamed as a broad issue, it is more likely that actors will consider it when discussing various forms of public policy. It is likely that when speaking about subsidies for fossil fuel companies in the 1980s, presidents would not mention the dangers of climate change, whilst today, climate change would likely be mentioned given the increased global attention.

While time and space might interact, here we are interested in their cumulative effects: when speaking in the 1990s to an audience that finds climate change pressing, certain transnational actors would likely mention it as a problem independently of the perceived pressingness of a climate breakdown at the time. Accordingly, certain transnational actors might even consider touching upon the issue while speaking at a business association today, because of its urgency.

@@

## 3.3 Operationalizing settings and pressingness

Our explanation for variation in problem-constructions is based on the setting of a speech and the pressingness of an aligned problem. This comprises our independent variables.

@@

*Pressingness*

Beyond setting, timing matters because issues aligned to certain problem-constructions vary in their pressingness over time. To capture the pressingness of problems, we use two proxies: lagged deforestation rates in the Brazilian Amazon and annual inflation rates. When deforestation rates are high, the problem of environmental conservation can be perceived as more pressing. Relatedly, when inflation is high, the problem of economic integration issues can be perceived as more pressing. We opt for lagging deforestation rates by one year as the consolidated rates for a certain year are published in the following year. Since both annual deforestation rates and annual inflation rates come in different measurements (i.e. percentage and kilometers square) we scale these values to facilitate interpretation of coefficients. Scaling variables do not affect statistical inference in regression models.

This approach to operationalizing pressingness also has its limitations. Ideally, we would like to add additional proxies of pressingness for social and sovereignty issues. However, social development indicators (e.g. life expectancy, education levels, access to sanitation) generally increased over time in Brazil. In terms of national sovereignty, the way we conceptualize the problem-construction entails internal and external threats which are not easily measured. Despite these limitations, we can draw some conclusions from the pressingness of environmental and economic issues and their aligned problem-constructions.

## 3.4 Inferential and descriptive strategies

@@

The dependent variables of the model are the share of speeches that mention the Amazon (model 1) and the share of a problem-construction (models 2-5). We chose to model each construction separately, and to focus on pure-types (i.e. statements that construct the Amazon as a single problem), to provide unambiguous interpretations and more robust coefficients. Our main explanatory variables are settings and pressignness, as described above.

@@

## 4.1 The Amazon as a transnational problem: setting and pressingness

We argue that presidents’ expectations about an audience at a certain setting and the pressingness of specific problems explain variations in when the Amazon becomes a  problem and how it is constructed as a type of problem it is. Our models, below, confirm the relationships expected by our framework.

@@

As we argue that the effects of setting and pressingness are cumulative, we start by first scrutinizing the relationship between the former and problem-construction. Model 1 addresses the question of when the Amazon is a problem to begin with. We model the probability of a speech generally mentioning the Amazon in relation to the reference category, Amazonian States within Brazil. The findings are intuitive: the likelihood of mentioning the Amazon in speeches decreases as presidents move away from the region and biome.

@@

We also argue that the pressingness of an aligned problem in time influences the likelihood of the president mentioning the Amazon and what they will say. We find evidence across our models that this holds. For instance, we find that presidents are less likely to generally mention the Amazon in speeches when deforestation rates increase and so the pressingness of environmental problems (model 1). The opposite is true for inflation. As inflation increases, and so the pressingness of economic problems, presidents are more likely to generally mention the  Amazon in speeches (model 1). This relationship is even more interesting when we look at how amazonian constructions vary in presidential speeches as environmental or economic problems become more pressing (models 2-3). We find strong evidence that when the pressingness of environmental problems increases, the probability of presidents constructing the Amazon as an environmental conservation problem decreases and the probability of constructing it as an issue of economic integration increases. Alternatively, as the pressingness of economic problems increases, the probability of presidents constructing the Amazon as an issue of economic integration decreases and the probability of constructing it as an issue of environmental conservation  increases.

In line with our framework, the pressingness of aligned problems relates to problem-constructions in intricate ways. By opting for an alternative problem-construction to that which is pressing, presidents likely side-step issues as they get more pressing, while boosting positive aligned outcomes to attempt and collect benefits as aligned issues become less pressing.

@@

Looking at speeches that mention the Amazon over time (the dependent variable of model 1) reviews several interesting trends. Figure 2, below, shows the proportion of speeches that mentions the Amazon in relation to all speeches in a year. By investigating its development in time, we can identify instances when the Amazon in itself becomes more pressing as a problem, and when not. On average, presidents mention the Amazon in 15.5% of their speeches ^[For comparison purposes, the averages for other policy objects in the same corpus of speeches are inequality (13.7%), criminality (17.3%), inflation (19.1%), and unemployment (13.3%).]. This suggests the Amazon is a topic of relative high importance for presidents generally. However, we observe four peak years (one standard deviation from the mean) in the figure: 1989, 1992, 2009, and 2019. We find that these peak years coincide with high-profile environmental events, but not with domestic policy adoption.

@@

Throughout the rest of the 1990s and 2000s, during the Cardoso (1995-2002) and Lula presidencies, there were no peaks. This is aligned with our model as the pressingness of the environmental problem was high during the period. In 2009, however, we observed another peak which coincided with the Copenhagen Summit. Lula led the delegation to the summit with a strong climate leadership self-image of “we deliver” [@franchini2019], in a moment when deforestation rates have been decreasing since 2005. From 2010 to 2016, except for 2015 when the Paris summit took place, we see a general decrease in mentions of the Amazon in all official presidential speeches. In this period,  the pressingness of environmental problems was increasing as deforestation rates were rising. The period was also marked by political and economic instability which culminated in the impeachment of President Rousseff in 2016. By 2016 mentions of Amazon in official speeches went down to 6%, their lowest share since 1985.

@@

The longitudinal trends in Amazonian speeches reveal a dynamic that our models did not capture: we find that all peak years relate to high-profile environmental events within the country that gained large repercussions within and outside of the country. These are events that received distinct coverage from transnational actors as media and civil society outside of Brazil, which is likely one of the reasons why it temporarily drives the topic up. This suggests that the Amazon is mentioned more in presidential speeches as transnational attention towards it increases. While our measure of pressingness of the environmental problem does not capture these events, these events change how pressing of a problem the Amazon is in certain years. In doing so, these events bring outside attention to the Amazon to which the president's response is more likely to be in the form of reaffirming Brazil’s sovereignty over the Amazon (model 5). That is, presidents tend to construct the Amazon as an issue of national sovereignty and side-step the aligned environmental conservation constructions around high-profile environmentally related events.

@@

While we identified the effects of setting and pressingness on problem-constructions (models 2-5), the developments of problem-constructions in time also reveal unique trends. In this section, we investigate how pure-types and mixed-types problem-constructions evolved over time. Figure 3, below, illustrates the annual share of pure-types problem-constructions.

@@

Pure economic integration constructions were the most frequent from 1985 to 2009. This is especially pertinent during Cardoso’s presidency (1995-2002), when construction of the Amazon as an issue of economic integration increased while environmental conservation constructions decreased. This was a period when the pressingness of the environmental problem grew as deforestation rates were increasing, so presidents were more likely to avoid environmental conservation constructions. When Lula took office (2003-2010), construction of the Amazon as an issue of environmental conservation and social development increased. During this period, the pressingness of environmental problems decreased along with deforestation rates, so presidents were more likely to employ environmental conservation constructions. From around 2010 until the mid-2010s, especially during the Rousseff (2011-2016) administration, we see a reversal at the top whereas environmental conservation and social development problem-constructions surpass economic integration. However, with the presidencies of Temer (2016-2018) and Bolsonaro (2019-2022), economic integration constructions become the most favored again.

@@

This paper investigated where, when, and how the Brazilian Amazon has been constructed as a problem in Brazilian presidential speeches since 1985. We have four main empirical findings. First, we find that presidents construct the Amazon as different problems depending on where they speak. When presidents speak outside the country and far away from the Amazon, they are more likely to construct the Amazon as an issue of environmental conservation. Within the country or the Amazonian region, presidents prefer constructing the Amazon as an issue of social development and economic integration. Second, we find that presidents are likely to construct the Amazon as an environmental conservation problem when environmental related issues are not pressing (e.g. low deforestation rates). When environmental related problems become more pressing (e.g. rising deforestation rates), presidents tend to side-step the issue and construct the Amazon as an issue of economic integration. Third, we find that high-profile environmental events, which receive strong international media attention, drive up the frequency at which the Amazon appears in discourse. Relevant Amazonian related policies, in turn, do not affect the discursive attention presidents pay to the Amazon. Finally, we find that we are now in a period when the Amazon as a binary of either environmental or an economic issue has been overcome, at least discursively, and Amazonian problem-constructions have become more mixed and balanced lately.

The conceptual implications of our findings are at least threefold. First, we provide an outlook of the region, its people and forest beyond its environmental nature by focusing on different problem-constructions. In so doing, we avoid assuming what type of problem the Amazon is and, instead, focus on how it has been represented. We show that investigating problem-representation to begin with, can highlight the extent to which certain perspectives of problems are assumed, ignored or privileged. Second, while many analyses of Brazilian environmental policies intend to dissect its national and international nature, they rarely adopt methodological strategies that distinguish how important actors behave differently in both scenarios (Hoschstetler and Keck, 2004). Our focus on presidential speeches affords us analytical leverage to identify transnational variation, as speeches take place in multiple settings over time. In doing so, we can contextualize problem-constructions in space and show how they vary. Third, we argue that another key factor in explaining variation in problem-construction is the pressingness of specific problems in time. As the environmental crisis increases in its salience, important actors are less likely to ignore it. They need to find strategies, such as side-stepping issues or boasting positive outcomes, in order to remain credible to their audiences.

>>>>

---
title: How has the Amazon been constructed as a problem? Presidential speeches and transnational politics since 1985
author:
- Livio Silva-Muller
- Henrique Sposito^[Both authors contributed equally to the article. Names are ordered alphabetically. Livio is a PhD Candidate in Sociology and affiliated to the Albert Hirschman Centre on Democracy. Henrique is a PhD Candidate in International Relations and Political Sciences and affiliated to the Centre for International Environmental Studies.]
date: 13 February 2023
output:
  pdf_document: default
  latex_engine: xelatex
  fig_caption: yes
  html_document:
    df_print: paged
fontsize: 12pt
mainfont: Times New Roman
header-includes:
   - \usepackage{floatrow}
   - \floatsetup[figure]{capposition=top}
   - \usepackage{float}
   - \floatplacement{figure}{H}
   - \usepackage[T1]{fontenc}
   - \usepackage[utf8]{inputenc}
   - |
    \providecommand{\keywords}[1]{\textbf{Keywords:} #1}
    \providecommand{\wordcount}[1]{\textbf{Word Count:}#1}
always_allow_html: TRUE
bibliography: references.bib
csl: chicago.csl
abstract: The Amazon is often approached as a transnational environmental problem without a discussion of how key actors represent it differently in space and time. We combine theories of problem-representation and transnationalism to analyze where, when, and how the Brazilian Amazon has been constructed as a problem. We develop a framework that explains variation in transnational problem-constructions based on speech setting and problem urgency To test our framework, we use supervised machine learning to classify statements about the Amazon in 6240 presidential speeches since 1985. We find that when presidents speak outside (or inside) the Amazon and environmental problems are less (or more) urgent, presidents are likelier to highlight issues of environmental conservation (or economic integration). Furthermore, we show how we are moving from a period when the Amazon as a binary of either environmental or an economic issue has been overcome, to a moment when diverse problem-constructions are deployed.

@@

```{=latex}
\wordcount{ 8190 (including abstract, text, references, and footnotes)}
```

@@

*We need to protect the Amazon from foreign interests. We need to exploit the Amazon's natural resources. We need to provide better living standards for the people in the Amazon. We need to preserve the Amazon as a standing ecosystem.* Each of these statements represent the Amazon as a particular problem: national sovereignty, economic integration, social development, and environmental conservation, respectively. These specific problems touch on common understandings of the Amazon as part of the larger socio-cultural history of the country. Brazilian governments have been portrayed in the literature as having held differing views on the Amazon, with some prioritizing national sovereignty and economic integration (military dictatorship and Bolsonaro), while others prioritize environmental conservation (Sarney and Lula). However, this oversimplifies government policies and discourse, which in reality are more intricate and contradictory.

By combining @bacchi1999’s theory of problem representation and @keck1998's theory of transnationalism, we place the way the objects of policies are constructed in discourse at the center of our analysis. We offer a framework that explains variation in discursive problem-constructions depending on speech setting and problem urgency as actors participate and shape transnational politics. We focus on speeches, rather than policies, because they allow us to capture the diversity of ways the Amazon, as a policy object that demands dedicated attention, has been discursively constructed as a problem. While many analyses of Brazilian environmental discourse and policies intend to dissect its national and international nature, they rarely adopt methodological strategies that distinguish how important actors behave differently in both scenarios [@hochstetler2007]. Our framework and methodology help identify and account for these variations. We show, for instance, that when speaking to farmers in Cuiaba, one of Brazil’s agricultural powerhouses, in a moment when the environmental problem is urgent, it is likely that any president would sidestep environmental issues, as President Lula did in the opening epigraph.

@@

We have four main findings. First, presidents adopt different constructions of the Amazon in different settings according to how they expect to leverage audiences for symbolic and/or  material gains. Second, presidents often focus on alternative problem-constructions when problems become more urgent; whereas they boast about positive outcomes when problems are less urgent. Third, high-profile environmental related events, which receive strong international media attention, increase the importance of the Amazon in presidential speeches. Alternatively, relevant Amazonian-related policies appear to not to attract much attention. Finally, while all presidents deploy multiple problem-constructions across space and over time, we are now in a period when the dictum between understanding the Amazon as an environment or as an economic issue has been overcome, at least discursively. Conceptually, our research shows how focusing on problem-constructions highlight what certain objects of policy are assumed to be, and when they are ignored or privileged, based on the transnational setting of the speech and the urgency of aligned problems.

@@

# 2. Theory: problem-constructions, transnational settings, and problem urgency

## 2.1 Problem-representation in presidential discourse

Social scientists often describe federal governments as proponents of a cohesive set of policies toward the Amazon [see @drummond2006; @padua2012; @franchini2019; @capobianco2019; @pereira2021]. The 1964 military dictatorship, for example, is associated with securing sovereignty in the region by populating it and integrating it into the national economy [@drummond2006]. Whereas, governments from the late-1980s up to the late-2000s are associated with a turn towards policies that focus on environmental conservation of the Amazon [@padua2012]. The presidencies of Rousseff (2011-2016) and Temer (2016-2018) are connected to the de-prioritizing of environmental conservation policies, while Bolsonaro (2019-2022) with the dismantling of environmental policies [@capobianco2019]. Although these works are important to understand how different governments acted towards the Amazon, they represent governments monolithically by bundling them into cohesive policy periods, or policy cycles, masking government diversity in policies and discourse.

@@

## 2.2 Transnational settings, urgency, and variation in problem-constructions

@@

Problem-constructions are choices made by presidents when speaking. If malleable identities across space and time are a defining feature of transnational politics, it follows that presidents' choice of problem-constructions depend on the president's views of an audience at a certain period. Specifically, how urgent a problem is in a certain space and/or at a certain time.

@@

Presidents construct problems and attempt to inform and/or leverage transnational audiences for symbolic or material gains according to their expectations of that audience. We conceptualize spaces in terms of settings. Settings are levels of aggregation for transnational audiences that make sense in light of the object of policy under study. This indicates that how presidents construct policy objects as problems can vary, for instance, within the country — speaking inside the Amazon region versus outside — because presidents identify these audiences as sources of distinct types of support and demand. The same holds true when speaking abroad — in Amazonian countries, non-Amazonian countries, and at international summits — where presidents’ expectations and identities of an audiences could diverge completely from that of domestic actors. In each one of these settings different problems might be more urgent and the president's expectations about what can be leveraged for material and symbolic gains might influence how a problem is implicitly or explicitly constructed.

@@

How presidents construct policy objects can also vary over time because the urgency of certain problems changes as new information arises. We conceptualize time as the perceived urgency of a broad and related issue to those surrounding a policy object. This indicates, for instance, that as climate change becomes more urgent and mainstreamed as a broad issue, it is more likely that actors will consider it when discussing various forms of public policy. It is likely that when speaking about subsidies for fossil fuel companies in the 1980s, presidents would not mention the dangers of climate change, whilst today, climate change would likely be mentioned given the increased global attention.

While time and space might interact, here we are interested in their cumulative effects: when speaking in the 1990s to an audience that finds climate change urgent, certain transnational actors would likely mention it as a problem independently of the perceived urgency of a climate breakdown at the time. Accordingly, certain transnational actors might even consider touching upon the issue while speaking at a business association today, because of its urgency.

@@

## 3.2 Operationalizing settings and urgency

Our explanation for variation in problem-constructions is based on the setting of a speech and the urgency of an aligned problem. This comprises our independent variables.

@@

*Urgency*

Beyond setting, timing matters because issues aligned to certain problem-constructions vary in their urgency over time. To capture the urgency of problems, we use two proxies: lagged deforestation rates in the Brazilian Amazon and annual inflation rates. When deforestation rates are high, the problem of environmental conservation can be perceived as more urgent. Relatedly, when inflation is high, the problem of economic integration issues can be perceived as more urgent. We opt for lagging deforestation rates by one year as the consolidated rates for a certain year are published in the following year. Since both annual deforestation rates and annual inflation rates come in different measurements (i.e. percentage and kilometers square) we scale these values to facilitate interpretation of coefficients. Scaling variables do not affect statistical inference in regression models.

This approach to operationalizing urgency also has its limitations. Ideally, we would like to add additional proxies of urgency for social and sovereignty issues. However, social development indicators (e.g. life expectancy, education levels, access to sanitation) generally increased over time in Brazil. In terms of national sovereignty, the way we conceptualize the problem-construction entails internal and external threats which are not easily measured. Despite these limitations, we can draw some conclusions from the urgency of environmental and economic issues and their aligned problem-constructions.

## 3.3 Inferential and descriptive strategies

@@

The dependent variables of the model are the share of speeches that mention the Amazon (model 1) and the share of a problem-construction (models 2-5). We chose to model each construction separately, and to focus on pure-types (i.e. statements that construct the Amazon as a single problem), to provide unambiguous interpretations and more robust coefficients. Our main explanatory variables are settings and urgency, as described above.

@@

## 4.1 The Amazon as a transnational problem: setting and urgency

We argue that presidents’ expectations about an audience at a certain setting and the urgency of specific problems explain variations in when the Amazon becomes a  problem and how it is constructed as a type of problem it is. Our models, below, confirm the relationships expected by our framework.

@@

As we argue that the effects of setting and urgency are cumulative, we start by first scrutinizing the relationship between the former and problem-construction. Model 1 addresses the question of when the Amazon is a problem to begin with. We model the probability of a speech generally mentioning the Amazon in relation to the reference category, Amazonian States within Brazil. The findings are intuitive: the likelihood of mentioning the Amazon in speeches decreases as presidents move away from the region and biome.

@@

We also argue that the urgency of an aligned problem in time influences the likelihood of the president mentioning the Amazon and what they will say. We find evidence across our models that this holds. For instance, we find that presidents are less likely to generally mention the Amazon in speeches when deforestation rates increase and so the urgency of environmental problems (model 1). The opposite is true for inflation. As inflation increases, and so the urgency of economic problems, presidents are more likely to generally mention the  Amazon in speeches (model 1). This relationship is even more interesting when we look at how amazonian constructions vary in presidential speeches as environmental or economic problems become more urgent (models 2-3). We find strong evidence that when the urgency of environmental problems increases, the probability of presidents constructing the Amazon as an environmental conservation problem decreases and the probability of constructing it as an issue of economic integration increases. Alternatively, as the urgency of economic problems increases, the probability of presidents constructing the Amazon as an issue of economic integration decreases and the probability of constructing it as an issue of environmental conservation  increases.

In line with our framework, the urgency of aligned problems relates to problem-constructions in intricate ways. By opting for an alternative problem-construction to that which is urgent, presidents likely side-step issues as they get more urgent, while boosting positive aligned outcomes to attempt and collect benefits as aligned issues become less urgent.

@@

Looking at speeches that mention the Amazon over time (the dependent variable of model 1) reviews several interesting trends. Figure 2, below, shows the proportion of speeches that mentions the Amazon in relation to all speeches in a year. By investigating its development in time, we can identify instances when the Amazon in itself becomes more urgent as a problem, and when not. On average, presidents mention the Amazon in 15.5% of their speeches ^[For comparison purposes, the averages for other policy objects in the same corpus of speeches are inequality (13.7%), criminality (17.3%), inflation (19.1%), and unemployment (13.3%).]. This suggests the Amazon is a topic of relative high importance for presidents generally. However, we observe four peak years (one standard deviation from the mean) in the figure: 1989, 1992, 2009, and 2019. We find that these peak years coincide with high-profile environmental events, but not with domestic policy adoption.

@@

Throughout the rest of the 1990s and 2000s, during the Cardoso (1995-2002) and Lula presidencies, there were no peaks. This is aligned with our model as the urgency of the environmental problem was high during the period. In 2009, however, we observed another peak which coincided with the Copenhagen Summit. Lula led the delegation to the summit with a strong climate leadership self-image of “we deliver” [@franchini2019], in a moment when deforestation rates have been decreasing since 2005. From 2010 to 2016, except for 2015 when the Paris summit took place, we see a general decrease in mentions of the Amazon in all official presidential speeches. In this period,  the urgency of environmental problems was increasing as deforestation rates were rising. The period was also marked by political and economic instability which culminated in the impeachment of President Rousseff in 2016. By 2016 mentions of Amazon in official speeches went down to 6%, their lowest share since 1985.

@@

The longitudinal trends in Amazonian speeches reveal a dynamic that our models did not capture: we find that all peak years relate to high-profile environmental events within the country that gained large repercussions within and outside of the country. These are events that received distinct coverage from transnational actors as media and civil society outside of Brazil, which is likely one of the reasons why it temporarily drives the topic up. This suggests that the Amazon is mentioned more in presidential speeches as transnational attention towards it increases. While our measure of urgency of the environmental problem does not capture these events, these events change how urgent of a problem the Amazon is in certain years. In doing so, these events bring outside attention to the Amazon to which the president's response is more likely to be in the form of reaffirming Brazil’s sovereignty over the Amazon (model 5). That is, presidents tend to construct the Amazon as an issue of national sovereignty and side-step the aligned environmental conservation constructions around high-profile environmentally related events.

@@

While we identified the effects of setting and urgency on problem-constructions (models 2-5), the developments of problem-constructions in time also reveal unique trends. In this section, we investigate how pure-types and mixed-types problem-constructions evolved over time. Figure 3, below, illustrates the annual share of pure-types problem-constructions.

@@

Pure economic integration constructions were the most frequent from 1985 to 2009. This is especially pertinent during Cardoso’s presidency (1995-2002), when construction of the Amazon as an issue of economic integration increased while environmental conservation constructions decreased. This was a period when the urgency of the environmental problem grew as deforestation rates were increasing, so presidents were more likely to avoid environmental conservation constructions. When Lula took office (2003-2010), construction of the Amazon as an issue of environmental conservation and social development increased. During this period, the urgency of environmental problems decreased along with deforestation rates, so presidents were more likely to employ environmental conservation constructions. From around 2010 until the mid-2010s, especially during the Rousseff (2011-2016) administration, we see a reversal at the top whereas environmental conservation and social development problem-constructions surpass economic integration. However, with the presidencies of Temer (2016-2018) and Bolsonaro (2019-2022), economic integration constructions become the most favored again.

@@

This paper investigated where, when, and how the Brazilian Amazon has been constructed as a problem in Brazilian presidential speeches since 1985. We have four main empirical findings. First, we find that presidents construct the Amazon as different problems depending on where they speak. When presidents speak outside the country and far away from the Amazon, they are more likely to construct the Amazon as an issue of environmental conservation. Within the country or the Amazonian region, presidents prefer constructing the Amazon as an issue of social development and economic integration. Second, we find that presidents are likely to construct the Amazon as an environmental conservation problem when environmental related issues are not urgent (e.g. low deforestation rates). When environmental related problems become more urgent (e.g. rising deforestation rates), presidents tend to side-step the issue and construct the Amazon as an issue of economic integration. Third, we find that high-profile environmental events, which receive strong international media attention, drive up the frequency at which the Amazon appears in discourse. Relevant Amazonian related policies, in turn, do not affect the discursive attention presidents pay to the Amazon. Finally, we find that we are now in a period when the Amazon as a binary of either environmental or an economic issue has been overcome, at least discursively, and Amazonian problem-constructions have become more mixed and balanced lately.

The conceptual implications of our findings are at least threefold. First, we provide an outlook of the region, its people and forest beyond its environmental nature by focusing on different problem-constructions. In so doing, we avoid assuming what type of problem the Amazon is and, instead, focus on how it has been represented. We show that investigating problem-representation to begin with, can highlight the extent to which certain perspectives of problems are assumed, ignored or privileged. Second, while many analyses of Brazilian environmental policies intend to dissect its national and international nature, they rarely adopt methodological strategies that distinguish how important actors behave differently in both scenarios (Hoschstetler and Keck, 2004). Our focus on presidential speeches affords us analytical leverage to identify transnational variation, as speeches take place in multiple settings over time. In doing so, we can contextualize problem-constructions in space and show how they vary. Third, we argue that another key factor in explaining variation in problem-construction is the urgency of specific problems in time. As the environmental crisis increases in its salience, important actors are less likely to ignore it. They need to find strategies, such as side-stepping issues or boasting positive outcomes, in order to remain credible to their audiences.
